Neural Circuit Organization of the Ctenophore Aboral Organ

### 2025-03-30 manuscript/Manuscript.qmd
ed main text

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. Each node was tagged <!--# is this true? not only a subset?  -->, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches <!--#  ref Burkhardt-->. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development<!--# ref -->. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral pole. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor <!--# what is this exactly? are there epithelial cells? -->of the aboral organ [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure.. -->. We could not detect specialized post-synaptic structures, in agreement with previous studies <!--# refs -->. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.

>>>>

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. Each node was tagged <!--# is this true? not only a subset?  -->, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure.. -->. We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.
